Detecting and quantifying parasite-induced host mortality from intensity data: method comparisons and limitations

### 2015-06-08 docs/parasite_mortality_manuscript.tex
Reformatted references for IJP

\textbf{Affiliations:} &
1) Department of Ecology, Evolution and Marine Biology, University of California,
Santa Barbara, Santa Barbara California \\

@@

\textbf{Keywords}: parasite aggregation, negative binomial distribution, Crofton Method, host survival function, $LD_{50}$

@@

@@

\begin{equation}
    h(\text{survival}; x, a, b) = \dfrac{e^{a - b \log(x)}}{1 + e^{a - b \log(x)}}
    \label{eq:logistic}
\end{equation}

@@

Unfortunately, this
flexibility in the distribution may also reduce our ability to detect PIHM. If
a negative binomial can be fit to the observed post-mortality parasite
distribution then, regardless of how lethal the parasite was, it will be
impossible to detect PIHM because there is no need for a more complex model.
Most observed parasite distributions are well fit by the negative binomial distribution \citep{Shaw1998}, suggesting
that systems where these methods are applicable may be more the exception than
the rule.  Furthermore, even when truncation of the negative binomial distribution is detected, it may be caused by other
processes such as within host density dependence, age dependent variation in host
resistance and/or heterogeneous infection rates \citep{McCallum2000a,Anderson1982a,Rousset1996}.  This means that in the event
that PIHM is detected, it may actually not be the result of PIHM.

@@

\singlespacing
\bibliographystyle{/Users/mqwilber/Dropbox/Documents/Bibformats/ecology_letters.bst}
\bibliography{/Users/mqwilber/Dropbox/Documents/Bibfiles/Projects_and_Permits-parasite_host_mortality}

@@

    \caption{A schematic representation of the iterative approach of the Crofton Method. \textbf{A)} The light gray shows the pre-mortality distribution that the Crofton Method is trying to estimate from the dark grey post-mortality distribution.  The Crofton Method proceeds by truncating the post-mortality data at different levels ($t_i$, e.g. $i= 0, \dots, 5$) and finding the parameters pre-mortality host population size ($N_p$), pre-mortality mean parasite intensity ($\mu_p$), and pre-mortality parasite aggregation ($k_p$) that best fit the truncated data. \textbf{B)} The parameter $N_p$ is then plotted against the truncation level $t_i$ to determine if a ``kink'' occurs in the parameter values  \citep{Lester1984}. This ``kink'' indicates that PIHM is occurring in the system. In the above example, PIHM is occurring in the system as visualized by the distinct ``kink'' at $t_3$.}
    \label{fig:crofton}

>>>>

\textbf{Affiliations:} &
1) Department of Ecology, Evolution and Marine Biology, University of California,
Santa Barbara, Santa Barbara California, United States of America \\

@@

\textbf{Keywords}: parasite aggregation, negative binomial distribution, Crofton Method, host survival function, lethal dose

@@

\bigskip

\noindent
\textbf{Highlights}
\begin{itemize}
    \item We develop and test a novel method for detecting and quantifying parasite-induced host mortality from parasite intensity data.
    \item Our method outperforms previous methods for detecting parasite-induced host mortality.
    \item Our method correctly quantifies parasite-induced host mortality in empirical data where mortality has been qualitatively reported
    \item  All current methods for detecting parasite-induced host mortality have fundamental and untestable assumptions that limit their applicability to many datasets.
\end{itemize}

@@

\begin{equation}
    h(\text{survival}; x, a, b) = \dfrac{\exp{(a - b \log(x))}}{1 + \exp{(a - b \log(x))}}
    \label{eq:logistic}
\end{equation}

@@

Unfortunately, this
flexibility in the distribution may also reduce our ability to detect PIHM. If
a negative binomial can be fit to the observed post-mortality parasite
distribution then, regardless of how lethal the parasite was, it will be
impossible to detect PIHM because there is no need for a more complex model.
Most observed parasite distributions are well fit by the negative binomial distribution \citep{Shaw1998}, suggesting
that systems where these methods are applicable may be more the exception than
the rule.  Furthermore, even when truncation of the negative binomial distribution is detected, it may be caused by other
processes such as within host density dependence, age dependent variation in host
resistance and/or heterogeneous infection rates \citep{Anderson1982a,Rousset1996, McCallum2000a}.  This means that in the event
that PIHM is detected, it may actually not be the result of PIHM.

@@

\singlespacing
\bibliographystyle{int_j_para.bst}
\bibliography{parasite_host_mortality}

@@

    \caption{A schematic representation of the iterative approach of the Crofton Method. \textbf{A)} The light gray shows the pre-mortality distribution that the Crofton Method is trying to estimate from the dark grey post-mortality distribution.  The Crofton Method proceeds by truncating the post-mortality data at different levels ($t_i$, e.g. $i= 0, \dots, 5$) and finding the pre-mortality host population size ($N_p$), pre-mortality mean parasite intensity ($\mu_p$), and pre-mortality parasite aggregation ($k_p$) that best fit the truncated data. \textbf{B)} The parameter $N_p$ is then plotted against the truncation level $t_i$ to determine if a ``kink'' occurs in the parameter values  \citep{Lester1984}. This ``kink'' indicates that PIHM is occurring in the system. In the above example, PIHM is occurring in the system as visualized by the distinct ``kink'' at $t_3$.}
    \label{fig:crofton}
